Almost Orthogonal Vectors

### 2014-12-11 project3/paper/paper.tex
edits

\title{Almost orthogonal vectors}

@@

\section{$n=2$, generic $N$}
In this section, we consider the problem of almost orthogonal vectors of dimensionality $2$.

@@

Observe that as before, we are interested in maximizing the square of the cosine of the angle between any two vectors in $v_1, v_2, \ldots, v_N$. Note that here, the angle between any two vectors in $v_j$ and $v_k$ in $v_1, v_2, \ldots, v_N$ can be expressed as $\sum_{i=j} ^{k-1} \theta_i$. Note that, however, if $\sum_{i=j}^{k-1} \theta_i \leq \pi/2$, then $\cos^2 \theta_j \geq \cos^2(\sum_{i=j}^{k-1} \theta_i)$ and that if $\sum_{i=j}^{k-1} \theta_i > \pi/2$, then $\cos^2 (\sum_{i=j}^{k-1} \theta_i) = \cos^2 (\pi - \sum_{i=j}^{k-1} \theta_i)$ which is less than $\cos^2 \theta_i$ for any $i$ in $$\{1, 2, \ldots, N-1, N\} \setminus \{j, j+1, \ldots, k-1\}$$, where $j, k \in \{1, 2, \ldots, N-1\}$ and $j \neq k$.

From this we can conclude that $\epsilon$ is equal to $\max_i \cos^2 \theta_i$. Then, if $j = \text{argmin } \theta_i$, then we see that $\epsilon = \cos^2 \theta_j$ by Lemma $2.2$.

@@

\section{A lower bound}

@@

\begin{lemma}
Let $A=(a_{i,j})$ be an $n$-by-$n$ real symmetric matrix with $a_{i,i}=1$ for all $i$ and $|a_{i,j}|\leq \sqrt{\epsilon}$ for all $i\neq j$. Let $d$ be the rank of $A$. Then
\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
\end{lemma}
\begin{proof}
We define $\lambda_1,\ldots, \lambda_n$ to be the eigenvalues of $A$; write $k$ as the number of nonzero eigenvalues. Their sum is the trace of $A$, equal to $n$. We know that the number of nonzero eigenvalues of a complex matrix is less than or equal to the rank of the matrix: so we have that $d\geq k$.

Recall the Cauchy-Schwarz for $\mathbb R^n$: given $(x_1,\ldots, x_n)$ and $(y_1,\ldots, y_n)\in \mathbb R^n$, we have
\[\left(\sum_{i=1}^nx_iy_i\right)^2\leq \left(\sum_{i=1}^n x_i^2\right)\left(\sum_{i=1}^ny_i^2\right).\]
In our case, let the $x_i$ equal $\lambda_i$, and let each $y_i$ equal 1 when $\lambda_i$ is nonzero and zero when $\lambda_i$ is zero. We get
\[n^2=\left(\sum_{i=1}^n \lambda_i\right)^2\leq \left(\sum_{i=1}^n \lambda_i^2\right) k\]
so that
\[\sum_{i=1}^n \lambda_i^2\geq\frac{n^2}{k}\geq \frac{n^2}{d}.\]
In fact, the sum $\sum_{i=1}^n \lambda_i^2$ is equal to the trace of $A^TA$, which we can compute explicitly to be $\sum_{i,j}a_{i,j}^2$. Hence we have that
\[\sum_{i,j}a_{i,j}^2\geq \frac{n^2}{d}.\]
But because $|a_{i,j}|\leq \sqrt\epsilon$ for $i\neq j$, we can bound the left hand side:
\[n+n(n-1)\epsilon\geq \left(\sum_{i=1}^j a_{i,i}^2\right)+\left(\sum_{i\neq j}a_{i,j}^2\right)=\sum_{i,j}a_{i,j}^2.\]
We obtain, therefore, that
\[n+n(n-1)\epsilon\geq \frac{n^2}{d},\]
and it clearly follows that
\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
\end{proof}
Using this lemma, we can now prove our lower bound on $\epsilon$.
\begin{theorem}
\label{lower_bound}
For any choice of $n$ and $N$, the minimum value of $\epsilon$ satisfies
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{theorem}
\begin{proof}
Pick any collection of $N$ unit vectors $v_1,\ldots, v_N$ in $\mathbb R^n$, and define $\epsilon$ as usual. Recall the definitions of $A$ and $G$:
\[A=\left(\begin{matrix}
| & | & \cdots & | & |\\
v_1 & v_2 & \cdots & v_{N-1} & v_N\\
| & | & \cdots & | & | \end{matrix}\right).\] 
\[G(v_1,\ldots, v_N)=A^TA=\left(\begin{matrix}
v_1\cdot v_1 & v_1\cdot v_2 & \ldots & v_1\cdot v_n\\
v_2\cdot v_2 & v_2\cdot v_2 & \ldots & v_2\cdot v_n\\
\vdots & \vdots & \ddots & \vdots\\
v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right).\]
Let $d$ be the rank of $G$. Since $A$ has rank at most $n$, so does $A^T$; since the image space of $G$ must be contained in the image space of $A^T$, we have that $d\leq n$. Note that $G$ is real and symmetric, with $1$s along the diagonal and all other entries with absolute value bounded by $\sqrt \epsilon$. We can apply the lemma to conclude
\[\epsilon\geq \frac{N-d}{d(N-1)}.\]
Since $d\leq n$, we have as desired
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{proof}
We present now a table of the bounds the theorem gives us for varying $n$ and $N$:
\begin{table}[h]
   \centering
    \begin{tabular}{ | c | c | c | c |}
    \hline
      & $n=2$ & $n=3$ & $n=4$ \\ \hline
    $N=2$ & $0$ & $-$ & $-$ \\ \hline
    $N=3$ & $1/4$ & $0$ & $-$ \\ \hline
    $N=4$ & $1/3$ & $1/9$ & $0$ \\ \hline
    $N=5$ & $3/8$ & $1/6$ & $1/16$ \\ \hline
    $N=6$ & $2/5$ & $1/5$ & $1/10$ \\ \hline
    $N=8$ &  $3/7$ & $5/21$ & $1/7$ \\ \hline
    $N=12$ & $5/11$ & $3/11$ & $2/11$ \\
    \hline
    \end{tabular}
    \caption {Lower bounds on $\epsilon$ for different values of $(n, N)$, given by Theorem \ref{lower_bound}.}
    \end{table}

@@

\section{Some optimal configurations}
Below we provide choices of $n$ and $N$ for which we can achieve the lower bound given by Theorem ~\ref{lower_bound}.
\subsection{$n=3,N=6$}
Let us consider collections of six vectors in $\mathbb R^3$. As can be seen in Table 1, the lower bound provided in this case by Theorem \ref{lower_bound} is $\frac 1 5$ as well. We shall now see, in fact, that this lower bound is attainable.
\begin{theorem}
For $n=3$ and $N=6$, the optimum value of $\epsilon$ is given by $\frac 1 5$.
\end{theorem}
\begin{proof}Consider the following 6 vectors on the unit sphere in $\R^3$ with variables $a$ and $b$. 
$$
(a,b,0),(a,0,b),(b,0,a)
$$
$$
(a,-b,0),(a,0,-b),(-b,0,a)
$$
There are three possible dot products we get from computing dot products between non-equal vectors chosen from these six vectors: $ab,-ab,a^2-b^2$. This gives us two possible values of $\epsilon$: $(a^2-b^2)^2,(ab)^2$. It would seem that an optimal configuration can be obtained by setting $a^2-b^2=ab$. If we add in the constraint that $a^2+b^2=1$ (since the six vectors we're interested in must live on the unit sphere), we get
$$
a=\sqrt{\frac{1}{10}(5-\sqrt{5}}, b=-\frac{1+\sqrt{5}}{2}\sqrt{\frac{1}{10}(5-\sqrt{5}}. 
$$
If we compute $\epsilon=(a^2-b^2)^2=(ab)^2$, we see that $\epsilon=\frac{1}{5}$. Coupled with our knowledge that $\frac 1 5$ is a lower bound on $\epsilon$, we conclude that these six points give us an optimal configuration.\end{proof}
\subsection{$N=n+1$}
Let us now consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space. From Theorem \ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so it suffices to show there exist some $n+1$ vectors in $n$-space that achieve an $\epsilon$ equal to $1/n^2$.

@@

Figure $5$ shows diagrams of a 2-dimensional simplex.

@@

@@

\section{A variation of the problem}

>>>>

\title{Almost Orthogonal Vectors}

@@

\section{$N$ vectors in $\R^2$}
In this section, we consider the problem of almost orthogonal vectors in $\R^2$.

@@

Observe that as before, we are interested in maximizing the square of the cosine of the angle between any two vectors in $v_1, v_2, \ldots, v_N$. Note that here, the angle between any two vectors in $v_j$ and $v_k$ in $v_1, v_2, \ldots, v_N$ can be expressed as $\sum_{i=j} ^{k-1} \theta_i$. Note that, however, if $\sum_{i=j}^{k-1} \theta_i \leq \pi/2$, then $\cos^2 \theta_j \geq \cos^2(\sum_{i=j}^{k-1} \theta_i)$ and that if $\sum_{i=j}^{k-1} \theta_i > \pi/2$, then $\cos^2 (\sum_{i=j}^{k-1} \theta_i) = \cos^2 (\pi - \sum_{i=j}^{k-1} \theta_i)$ which is less than $\cos^2 \theta_i$ for any $i$ in $$\{1, 2, \ldots, N-1, N\} \setminus \{j, j+1, \ldots, k-1\},$$ where $j, k \in \{1, 2, \ldots, N-1\}$ and $j \neq k$.

From this we can conclude that $\epsilon$ is equal to $\max_i \cos^2 \theta_i$. Then, if $j = \text{argmin } \theta_i$, we see that $\epsilon = \cos^2 \theta_j$ by Lemma $2.2$.

@@

\section{A lower bound on $\epsilon$}

@@

\begin{lemma}
Let $A=(a_{i,j})$ be an $n\times n$ real symmetric matrix with $a_{i,i}=1$ for all $i$ and $|a_{i,j}|\leq \sqrt{\epsilon}$ for all $i\neq j$. Let $d$ be the rank of $A$. Then
\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
\end{lemma}
\begin{proof}
We define $\lambda_1,\ldots, \lambda_n$ to be the eigenvalues of $A$; write $k$ as the number of nonzero eigenvalues. Their sum is the trace of $A$, equal to $n$. We know that the number of nonzero eigenvalues of a complex matrix is less than or equal to the rank of the matrix: so we have that $d\geq k$.

Recall the Cauchy-Schwarz for $\mathbb R^n$: given $(x_1,\ldots, x_n)$ and $(y_1,\ldots, y_n)\in \mathbb R^n$, we have
\[\left(\sum_{i=1}^nx_iy_i\right)^2\leq \left(\sum_{i=1}^n x_i^2\right)\left(\sum_{i=1}^ny_i^2\right).\]
In our case, let the $x_i$ equal $\lambda_i$, and let each $y_i$ equal 1 when $\lambda_i$ is nonzero and zero when $\lambda_i$ is zero. We get
\[n^2=\left(\sum_{i=1}^n \lambda_i\right)^2\leq \left(\sum_{i=1}^n \lambda_i^2\right) k\]
so that
\[\sum_{i=1}^n \lambda_i^2\geq\frac{n^2}{k}\geq \frac{n^2}{d}.\]
In fact, the sum $\sum_{i=1}^n \lambda_i^2$ is equal to the trace of $A^TA$, which we can compute explicitly to be $\sum_{i,j}a_{i,j}^2$. Hence we have that
\[\sum_{i,j}a_{i,j}^2\geq \frac{n^2}{d}.\]
But because $|a_{i,j}|\leq \sqrt\epsilon$ for $i\neq j$, we can bound the left hand side:
\[n+n(n-1)\epsilon\geq \left(\sum_{i=1}^j a_{i,i}^2\right)+\left(\sum_{i\neq j}a_{i,j}^2\right)=\sum_{i,j}a_{i,j}^2.\]
We obtain, therefore, that
\[n+n(n-1)\epsilon\geq \frac{n^2}{d},\]
and it clearly follows that
\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
\end{proof}
Using this lemma, we can now prove our lower bound on $\epsilon$.
\begin{theorem}
\label{lower_bound}
For any choice of $n$ and $N$, the minimum value of $\epsilon$ satisfies
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{theorem}
\begin{proof}
Pick any collection of $N$ unit vectors $v_1,\ldots, v_N$ in $\mathbb R^n$, and define $\epsilon$ as usual. Recall the definitions of $A$ and $G$:
\[A=\left(\begin{matrix}
| & | & \cdots & | & |\\
v_1 & v_2 & \cdots & v_{N-1} & v_N\\
| & | & \cdots & | & | \end{matrix}\right)\] 
\[G(v_1,\ldots, v_N)=A^TA=\left(\begin{matrix}
v_1\cdot v_1 & v_1\cdot v_2 & \ldots & v_1\cdot v_n\\
v_2\cdot v_2 & v_2\cdot v_2 & \ldots & v_2\cdot v_n\\
\vdots & \vdots & \ddots & \vdots\\
v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right)\]
Let $d$ be the rank of $G$. Since $A$ has rank at most $n$, so does $A^T$; since the image space of $G$ must be contained in the image space of $A^T$, we have that $d\leq n$. Note that $G$ is real and symmetric, with $1$s along the diagonal and all other entries with absolute value bounded by $\sqrt \epsilon$. We can apply the lemma to conclude
\[\epsilon\geq \frac{N-d}{d(N-1)}.\]
Since $d\leq n$, we have as desired
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{proof}
We present now a table of the bounds the theorem gives us for varying $n$ and $N$:
\begin{table}[h]
   \centering
    \begin{tabular}{ | c | c | c | c |}
    \hline
      & $n=2$ & $n=3$ & $n=4$ \\ \hline
    $N=2$ & $0$ & $-$ & $-$ \\ \hline
    $N=3$ & $1/4$ & $0$ & $-$ \\ \hline
    $N=4$ & $1/3$ & $1/9$ & $0$ \\ \hline
    $N=5$ & $3/8$ & $1/6$ & $1/16$ \\ \hline
    $N=6$ & $2/5$ & $1/5$ & $1/10$ \\ \hline
    $N=8$ &  $3/7$ & $5/21$ & $1/7$ \\ \hline
    $N=12$ & $5/11$ & $3/11$ & $2/11$ \\
    \hline
    \end{tabular}
    \caption {Lower bounds on $\epsilon$ for different values of $(n, N)$, given by Theorem \ref{lower_bound}.}
    \end{table}

@@

\section{Some optimal configurations}
Below we provide choices of $n$ and $N$ for which we can achieve the lower bound given by Theorem ~\ref{lower_bound}.
\subsection{$n=3,N=6$}
Let us consider collections of six vectors in $\mathbb R^3$. As can be seen in Table 1, the lower bound provided in this case by Theorem \ref{lower_bound} is $\frac 1 5$ as well. We shall now see, in fact, that this lower bound is attainable.
\begin{theorem}
For $n=3$ and $N=6$, the optimum value of $\epsilon$ is given by $\frac 1 5$.
\end{theorem}
\begin{proof}Consider the following 6 vectors on the unit sphere in $\R^3$ with variables $a$ and $b$. 
$$
(a,b,0),(a,0,b),(b,0,a)
$$
$$
(a,-b,0),(a,0,-b),(-b,0,a)
$$
There are three possible dot products we get from computing dot products between non-equal vectors chosen from these six vectors: $ab,-ab,a^2-b^2$. This gives us two possible values of $\epsilon$: $(a^2-b^2)^2,(ab)^2$. It would seem that an optimal configuration can be obtained by setting $a^2-b^2=ab$. If we add in the constraint that $a^2+b^2=1$ (since the six vectors we're interested in must live on the unit sphere), we get
$$
a=\sqrt{\frac{1}{10}(5-\sqrt{5}}$$$$ b=-\frac{1+\sqrt{5}}{2}\sqrt{\frac{1}{10}(5-\sqrt{5}}. 
$$
If we compute $\epsilon=(a^2-b^2)^2=(ab)^2$, we see that $\epsilon=\frac{1}{5}$. Coupled with our knowledge that $\frac 1 5$ is a lower bound on $\epsilon$, we conclude that these six points give us an optimal configuration.\end{proof}
\subsection{$N=n+1$}
Let us now consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space. From Theorem \ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so it suffices to show there exist some $n+1$ vectors in $n$-space that achieve an $\epsilon$ equal to $1/n^2$.

@@

Figure $5$ shows diagrams of a 2-dimensional simplex. Notice that an $n$ dimensional simplex is constructed in $n+1$ dimensions, but exists in a hyperplane of $\R^{n+1}$.

@@

As another note, in some instances the algorithm sometimes finds extrema that are not the extrema we have identified in cases where we provide optimal configurations. For example, for $n=3,N=4$, the $\epsilon$ value attained is very close to the optimal epsilon value of $\frac{1}{9}$, however the corresponding configuration is not a tetrahedron. This implies that in some cases there may be multiple configurations of vectors that are inequivalent up to rotation yet produce the optimal $\epsilon$ value.

@@

\section{A variation}
